---
title: "Motor Trend Magazine mpg Transmission Analysis"
author: "Ann Crawford"
date: "October 2, 2017"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

#Executive Summary
This paper uses regression models and exploratory data analysis on the mtcars dataset to examine transmision type (mannual or automatic) as a predictor for miles per gallon. 
Specifically address these questions"" 
1.  Is an automatic or manual transmission better for MPG? In this case better means higher value for miles per gallon.  
2.  Quantify the MPG difference between automatic and manual transmissions.   
 
## Steps to Analysis
1. Examine the DataSet
2. Analyze a simple linear regression mpg to transmission: automatic vs manual 
3. Review other perdictors using multivariate analysis to quantify transmission as a predictor. 

## Conclusion
The strongest indication of mpg is defined by the inverse realation of the weight to mpg. Transmission and weight are correlated. Including transmission in regression analysis add little information for explaining the variance for mpg. 

### Examine the DataSet
The mtcars data frame contains 32 observations on 11 variables. vs,am, gear, carb and cyl can be treated as factors. For details consult help **?mtcars**
```{r cars}

y <- mtcars$mpg
x <- mtcars$am
# check for nas 
sapply(mtcars,function(x) sum(is.na(x)))
sapply(mtcars, function(x) length(unique(x)))
```
```{r slr,echo=FALSE}
fit1 <- lm(mpg ~ am , data= mtcars)
summary(fit1)

e <- resid(fit1)
```
### Analyze a simple linear regression mpg to transmission: automatic vs manual 
Since question 1 specifcally asks to compare mpg to tranmission, the box plot shows that mannual transmissions have higher mpg in general with some variance. To further explore the variance and the value of am as a predictor, start with simple linear regression using transmisison as the single perdictor for mpg and review the residuals.
fit1 <- lm(mpg ~ am , data= mtcars)
e <- resid(fit1)
The model fit returns a pvalue of `r summary(fit1)$coefficients[2,4]` indicating statical significance at the 95% level. R squared  = `r  summary(fit1)$r.squared` indicating transmission accounts for roughly 36% of the variablity in mpg.

```{r plottransMPG, echo=FALSE, warning=FALSE}
library(ggplot2)
g = ggplot(mtcars, aes(x=factor(am), y=mpg)) + geom_boxplot()
g = g + ylab(" Miles/(US) gallon")
g = g + xlab("Transmission (0 = automatic, 1 = manual)")
g
```


### Residual Plot shows errors with the model design
The residual plot shows, the relationship is **not a simple linear regression**. The pattern in residual plot reveals errors with the model design. Residuals represent variation left unexplained by our model. Points should be radomly clustered around the 0.
```{r slr2plot, ,echo=FALSE}
plot(mtcars$mpg, e,
     ylab = "Residuals", xlab = "Miles/gallon(US)",
     main = "Mpg~am Residual Plot")
abline(0,0)

g2 = ggplot(mtcars, aes(x = am, y = mpg))
g2 = g2 + ylab(" Miles/(US) gallon")
g2 = g2 + xlab("Transmission (0 = automatic, 1 = manual)")
g2 = g2 + geom_point(size = 7, colour = "black", alpha=0.5)
g2 = g2 + geom_point(size = 5, colour = "blue", alpha=0.2)
g2 = g2 + geom_smooth(method = "lm", colour = "black")
g2
```

# Multvariate Analysis

This section use multi-variate linear regression to explore how the other variables in the mtcars dataset may be related to mpg.

First fit a model with all the variables. But the results contradict the initial findings, where the p values show no statisical significance at the 95% level, all greater than 0.05, with wt having the lowest p value. This indicates and interation between the variables. 

## Checking the correlation of perdictors 
The correction matrix confirms the amount of correlation in the mtcars varaibles. It seems that wt is a strong predictor for the mpg response variable. 

Ploting weight mpg and color coding the transmission shows a strong inverse relationship between wt and am.
cor(mtcars[,1:11])

```{r multiVariate}
fit2 <- lm(mpg ~ ., data = mtcars)
summary(fit2)

```
## Multiple Linear regression 

```{r multiplot, echo=FALSE}
# Plot 

#plot(wt, mpg,data=mtcars, main="Weight and mpg", 
#  	xlab="Car Weight ", ylab="Miles Per Gallon ", pch=19)
#fitwt  <- lm(mpg~wt,mtcars)
#abline(fitwt)

cormat <- cor(mtcars)



fitwtam <- lm(mpg ~ wt*am, data = mtcars)

Predicted  <- predict(fitwtam, mtcars)

mydf = cbind(mtcars, Predicted)

qplot(mtcars$wt, mtcars$mpg, colour = as.factor(am), main = "mpg vs Wt, Trans (Manual=1, Automatic=0)",
   data = mtcars)

```
# Is weight or weight as an interaction variable with transsmission a better model for predicting mpg?
Given the correlation betwwen wt and mpg is ??

```{r wtvsam}

fitwtam <- lm(mpg~am*wt, mtcars)
summary(fitwtam)

fitwt <- lm(mpg~wt, mtcars)
summary(fitwt)

```